---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set( warning=FALSE, message=FALSE)
```

## Loading and preprocessing the data
The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

- date: The date on which the measurement was taken in YYYY-MM-DD format

- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset. The code for this analysis assumes the dataset is in a file named "activity.csv" in the data subdirectory of the working
directory. 

The data is read in using read.csv, and the date is parsed using lubridate. 

```{r load data}
library(lubridate)
data <-read.csv("./data/activity.csv")
data <- transform(data, date = ymd(as.character(date)))

```




## What is mean total number of steps taken per day?


```{r mean and median daily steps}
## group by day and calculate total steps per day, and the mean and
## median of the total steps across all the days.
library(dplyr)
data_by_date <- group_by(data, date)
total_steps <- summarise(data_by_date, total_steps = sum(steps, na.rm = TRUE))
plot(total_steps, type = "h", xlab = "Date", ylab = "Total Steps")
mean_steps <- mean(total_steps$total_steps)
median_steps <- median(total_steps$total_steps)
print(c("The mean total steps per day is ", mean_steps), quote = FALSE)
print(c("The median total steps per day is ", median_steps), quote = FALSE)

```


## What is the average daily activity pattern?
```{r daily activity}
# group by interval and average the number of steps per interval across the
# days. 
data_by_int <- group_by(data, interval)
interval_avg <- summarise(data_by_int, mean = mean(steps, na.rm = TRUE))
plot(interval_avg, type = "l", ylab = "Mean steps", xlab = "Interval")

# calculate the average number of steps taken across the days
# during each interval. Add it to the dataset in a column called 
# "abisteps", which means average by interval steps. 
# Calcuate which interval has the highest mean steps. 
data <- transform(data, interval = as.factor(interval))
data <- transform(data, abisteps = ave(data$steps, interval,
                                       FUN = function(x) mean (x, na.rm = TRUE)))
most_active <- interval_avg[which(interval_avg$mean == 
                                    max(interval_avg$mean)),1]
print(c("The interval with the highest average number of steps is ", most_active),
      quote = FALSE)
```





## Imputing missing values

```{r imputing missing values}
# calulate and report the number of missing values in the dataset.
print(c("The number of missing values in the dataset is ",sum(is.na(data$steps))),
        quote = FALSE)
#create an interpolated dataset where NA values in the steps column
#are replaced with the average number of steps taken during that interval.
interpolated_data <- transform(data, steps = ifelse(is.na(steps), 
                                                    abisteps, steps))

```
## Are there differences in activity patterns between weekdays and weekends?
```{r weekday and weekend patterns}

```