---
title: "PeerAssessment1"
author: "putsif"
date: "Tuesday, May 12, 2015"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set( warning=FALSE, message=FALSE)
```


###Loading the data 
The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

- date: The date on which the measurement was taken in YYYY-MM-DD format

- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset. The code for this analysis assumes the dataset is present in the working directory in a file called "activity.csv". 

The data is read in using read.csv, and the date is parsed using lubridate. 

```{r load data}
library(lubridate)
data <-read.csv("./data/activity.csv")
data <- transform(data, date = ymd(as.character(date)))

```

Using the dplyr package, group the data by the date and calculate the total steps taken each day. Then calculate the mean and median of the daily step totals.  

```{r calculate total steps}
library(dplyr)
data <- group_by(data, date)
total_steps <- summarise(data, sum(steps, na.rm = TRUE))
mean_steps <- mean(total_steps$total_steps)
median_steps <- median(total_steps$total_steps)

```

